Structure Learning for $VAR(p)$ Models

### 2019-03-06 paper.tex
Added some definitions worked on lemma.

\def\gc{\overset{\text{GC}}{\rightarrow}}
\def\te{\overset{\mathcal{T}}{\rightarrow}}

@@

\section{Theory}
\label{sec:theory}
Consider a zero mean wide sense stationary, vector valued and discrete
time stochastic process $x(t) \in L^n_2(\Omega);\ t \in \Z$, having the matrix
valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$ and spectra $S(\omega)$:

@@

\begin{definition}[Granger Causality]
  For the WSS series $x(t)$ we say that component $x_j$
  \textit{Granger-Causes} (GC) component $x_i$ (with respect to
  $x(t)$) and write $x_j \gc x_i$ if given Hilbert spaces
  $\mathcal{H}^{(i)}_{t - 1}$, $\mathcal{H}^{(i, j)}_{t - 1}$
  generated by $\{x_i(t - \tau)\}_{\tau = 1}^\infty$ and
  $\{x_i(t - \tau), x_j(t - s\}_{\tau, s = 1}^\infty$ respectively we have

\begin{equation}
  \xi[x_i(t) \ |\ \mathcal{H}^{(i)}_{t - 1}] > \xi[x_i(t) \ |\ \mathcal{H}^{(i, j)}_{t - 1}],
\end{equation}

where
$\xi[x \ |\ \mathcal{H}] = \E (x - \hat{\E}[x \ |\ \mathcal{H}])^2$ and
$\hat{\E}[x \ |\ \mathcal{H}] = \text{proj}_{\mathcal{H}}(x)$ denotes
the (unique) projection onto the Hilbert space $\mathcal{H}$.
\end{definition}

@@

\begin{equation}
\label{eqn:wold}
  x(t) = c(t) + \sum_{\tau = 0}^\infty A(\tau) z(t - \tau);\ z(t) \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, I),
\end{equation}

where $c(t)$ is a perfectly predictable sequence\footnote{$c(t)$ is
  perfectly predictable if any sample $c(t_0)$ is enough to determine
  $c(t)$ for every $t$.  For example,
  $c(t) = \text{sin}(2\pi t + \Theta);\; \Theta \sim \mathcal{U}[-\pi, \pi]$}, and
$z(t) \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, I)$.  We will assume
that $c(t)$ can be effectively estimated and removed and take
$c(t) = 0$.  Furthermore, we assume that the spectra is uniformly
bounded away from $0$:
$\exists \delta:\ \delta I \preceq S(\omega) \preceq \frac{1}{\delta} I\ \forall \omega$ so that the Wold
representation can be inverted to yield the $VAR(\infty)$ form

\begin{equation}
  x(t) = \sum_{\tau = 1}^\infty B(\tau) x(t - \tau) + v(t);\ v(t) \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, \Sigma_v).
\end{equation}

\begin{theorem}[Granger Causality Equivalences]
  Given W.S.S. vector valued process $x(t)$ the following are equivalent:

  \begin{enumerate}
    \item{$x_j \gc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$}
    \item{$\mathcal{H}_t^{(i)} \perp \mathcal{H}_{t - 1}^{(j)}\ |\ \mathcal{H}_{t - 1}^{(i)}$}
    \item{$\mathcal{F}_{j \rightarrow i} \defeq \ln\frac{\xi_i}{\xi_{ij}} > 0$}
    \item{\hl{Wold $A(\tau)$ condition}}
    \item{Tie in the 0s of $S(\omega)^{-1}$, though this isn't immediately directed.}
  \end{enumerate}
\end{theorem}

@@

\begin{theorem}[Invariance Under Invertible and Deterministic Modifications]
  Let $F(z), G(z)$ describe univariate and invertible
  linear-time-invariant filters.  And, let $f(t), g(t)$ be perfectly
  predictable processes.  Then,

  \begin{equation}
    x_j \gc x_i \iff F(x_j - f)(t) \gc G(x_i - g)(t)
  \end{equation}
\end{theorem}

@@

\begin{theorem}[Tree Recovery]
  \label{thm:tree_recovery}
  If the Granger-Causality graph $\mathcal{G}$ for $x(t)$ is a
  directed forest, then $\mathcal{G}$ can be inferred from pairwise
  causality tests alone.
\end{theorem}

\section{Structure Learning}
\subsection{Edge-Wise Grouped LASSO}
\subsection{Bayesian Posterior Thresholding}
Consider the Bayesian model

>>>>

\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger-causality graph
\def\VAR{\mathsf{VAR}}  % VAR(p) model

\newcommand{\linE}[2]{\hat{\E}[#1\ |\ #2]}  % Linear projection
\newcommand{\linEerr}[2]{\xi[#1\ |\ #2]}  % Error of linear projection
\newcommand{\pa}[1]{P(#1)}  % Parents of a node
\newcommand{\anc}[1]{\mathcal{A}(#1)}  % Ancestors of a node

@@

\section{Theory}
\label{sec:theory}
Consider a zero mean wide sense stationary, vector valued and discrete
time stochastic process $x(t) \in L^n_2(\Omega);\ t \in \Z$, having the matrix
valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$, which we assume to be
absolutely summable, and spectra $S(\omega)$:

@@

\begin{definition}[Granger Causality]
  For the WSS series $x(t)$ we say that component $x_j$
  \textit{Granger-Causes} (GC) component $x_i$ (with respect to $x$)
  and write $x_j \gc x_i$ if given Hilbert spaces
  $\mathcal{H}_{t - 1}$, $\mathcal{H}^{-j}_{t - 1}$
  generated by $\{x_i(t - \tau)\}_{(\tau = 1, i = 1)}^{(\infty, n)}$ and
  $\{x_i(t - \tau)\}_{(\tau = 1, i \ne j)}^{(\infty, n)}$ respectively we have

\begin{equation}
  \linEerr{x_i(t)}{\mathcal{H}_{t - 1}} < \linEerr{x_i(t)}{\mathcal{H}^{-j}_{t - 1}},
\end{equation}

where
$\xi[x \ |\ \mathcal{H}] = \E (x - \linE{x}{\mathcal{H}})^2$ and
$\linE{x}{\mathcal{H}} = \text{proj}_{\mathcal{H}}(x)$ denotes
the (unique) projection onto the Hilbert space $\mathcal{H}$.
\end{definition}

@@

\begin{equation}
\label{eqn:wold}
  x(t) = c(t) + \sum_{\tau = 0}^\infty A(\tau) z(t - \tau),
\end{equation}

where $c(t)$ is a perfectly predictable sequence\footnote{$c(t)$ is
  perfectly predictable if any sample $c(t_0)$ is enough to determine
  $c(t)$ for every $t$.  For example,
  $c(t) = \text{sin}(2\pi t + \Theta);\; \Theta \sim \mathcal{U}[-\pi, \pi]$}, and
$z(t)$ is an uncorrelated sequence having covariance $I$.  We will assume
that $c(t)$ can be effectively estimated and removed and take
$c(t) = 0$.  Furthermore, we assume that the spectra is uniformly
bounded away from $0$:
$\exists \delta:\ \delta I \preceq S(\omega) \preceq \frac{1}{\delta} I\ \forall \omega$ so that the Wold
representation can be inverted to yield the $VAR(\infty)$ form

\begin{equation}
  \label{eqn:ar_representation}
  x(t) = \sum_{\tau = 1}^\infty B(\tau) x(t - \tau) + v(t),
\end{equation}

where $v(t)$ is an uncorrelated sequence with covariance $\Sigma_v$.

\begin{theorem}[Granger Causality Equivalences]
  Let $x(t)$ be a WSS process with absolutely summable covariance
  sequence, and uniformly bounded spectral density.  Denote by
  $\xi_{ij} \defeq \linEerr{x_i(t)}{\mathcal{H}^{-j}}$ and
  $\xi_i \defeq \linEerr{x_i(t)}{\mathcal{H}_t}$.  Then, the following are equivalent:

  \begin{enumerate}
    \item{$x_j \not\gc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$}
    \item{$\mathcal{F}_{j \rightarrow i} \defeq \ln\frac{\xi_i}{\xi_{ij}} = 0$}
    \item{$\mathcal{H}_t^{i} \perp \mathcal{H}_{t - 1}^{j}\ |\ \mathcal{H}_{t - 1}^{-j}$}
    \item{\hl{Wold $A(\tau)$ condition}}
    \item{Tie in the 0s of $S(\omega)^{-1}$, though this isn't immediately directed.}
  \end{enumerate}
\end{theorem}

The equivalence $(1) \iff (3)$ is essentially a restatement of the
definition, but is in line with the seminal work of Geweke
\cite{geweke1982}, \cite{geweke1984}.  Statement (4) means that
$(y - \linE{y}{\mathcal{H}_{t - 1}^{-j}}) \perp (z -
\linE{z}{\mathcal{H}_{t - 1}^{-j}})\ \forall y \in \mathcal{H}_t^i,\ z \in
\mathcal{H}_{t - 1}^j$, which is an appealing geometric statement that
there is no direct feedback from $x_j$ to $x_i$ \cite{lindquist}.

\begin{proof}
$(1) \implies (2)$: The projection for $x_i(t)$ onto $\mathcal{H}_{t - 1}$ is (by definition) given by a form similar to equation \ref{eqn:ar_representation}:

\[
  \E|x_i(t) - \linE{x_i(t)}{\mathcal{H}_{t - 1}}|^2 = \E|v_i(t) + \sum_{\tau = 1}^\infty \sum_{k = 1}^n (B_{i, k}(\tau) - \hat{B}_{i, k}(\tau))x_k(t - \tau)|^2.
\]

 Since $v(t)$ in equation \ref{eqn:ar_representation} is temporally uncorrelated, it follows that the optimal projection is given by the model coefficients themselves.  This holds similarly for the projection onto $\mathcal{H}_{t - 1}^{-j}$.  Then by $(1)$ we have

\[
  \E|x_i(t) - \sum_{\tau = 1}^\infty \sum_{k = 1}^n B_{i, k}(\tau)x_k(t - \tau)|^2 = \E|x_i(t) - \sum_{\tau = 1}^\infty\sum_{k \ne j}B_{i, k}(\tau)x_k(t - \tau)|^2
\]

By the uniqueness of the projection we must have $\forall \tau\  B_{i, j}(\tau) = 0$.

$(2) \implies (4)$: In computing $(y - \linE{y}{\mathcal{H}_{t = 1}^{-j}})$ for $y \in \mathcal{H}_t^i$ it is sufficient to consider $y = x_i(t)$ since $\mathcal{H}_{t - 1}^i \subseteq \mathcal{H}_{t - 1}^{-j}$, in which case $(x_i(t) - \linE{x_i(t)}{\mathcal{H}_{t = 1}^{-j}}) = v_i(t)$.  $(4)$ follows since $v_i(t) \perp \mathcal{H}_{t - 1}$ and $\forall z \in \mathcal{H}_{t - 1}^j\ (z - \linE{z}{\mathcal{H}_{t - 1}^{-j}}) \in \mathcal{H}_{t - 1}$

\end{proof}

@@

@@

\begin{corollary}[Invariance Under Invertible and Deterministic Modifications]
  Let $F(z), G(z)$ describe univariate and invertible
  linear-time-invariant filters.  And, let $f(t), g(t)$ be perfectly
  predictable processes.  Then,

  \begin{equation}
    x_j \gc x_i \iff F(x_j - f)(t) \gc G(x_i - g)(t)
  \end{equation}
\end{corollary}

\subsection{Pairwise Granger Causality}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  That
is, $x_k$ may be a common ancestor that when observed completely
explains the connection from $x_j$ to $x_i$.  In particular, we will
study \textit{pairwise} Granger-causality, where $x_i \pwgc x_j$ if
$x_i$ Granger-causes $x_j$ with respect only to $(x_i, x_j)$.

We establish first some notation.  Remaining in the context of a
collection of $n$ WSS processes
$x(t) = \big(x_1(t), \ldots, x_n(t)\big)$, we define the Granger-causality
graph $\gcg$ to be the directed graph formed on $n$ vertices where an
edge $(j, i) \in \gcg$ if and only if $x_j \gc x_i$.  We refer to the
set of parents of $x_i$ in $\gcg$ by $\pa{i}$, that is,
$j \in \pa{i} \iff x_j \gc x_i$.  The set of \textit{ancestors} of
$x_i$ in $\gcg$ will be denoted $\anc{i}$, that is, $j \in \anc{i}$ if
and only if there is a directed path in $\gcg$ from $j$ to $i$.  Even
though the definitions of Granger-causality naturally suggests that
each node in the graph should have a self-loop (unless the node is
simply independent noise) we will not include nodes in their own
parent sets: $i \not\in \pa{i} \subseteq \anc{i}$.  We note that $\gcg$ need not be a DAG.

The edges of the graph can be given a general notion of ``weight'' by
associating an edge $(j, i)$ with an LTI filter
$\{B_{ij}(\tau)\}_{\tau = 1}^\infty$ from the $\VAR$ representation of equation
\ref{eqn:ar_representation}.

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
% I'm actually not sure this is true, I think that if there is a common ancestor it is
% ambiguous.  This may be true for transfer entropy, but for Granger-causality even if
% the information arriving at i is later than to j, it may be less corrupt and hence
% still facilitate better predictions about k's future.

Finally, the \textit{propagation delay} from $j$ to $i$, denoted $d(j, i)$, will be defined as

\[
  d(j, i) = \text{min}\big\{\sum_{k = 0}^{K - 1} d(p_k, p_{k + 1}})\ |\ i = p_0 \rightarrow p_1 \rightarrow \cdots \rightarrow p_K = j \text{ is a path in } \gcg \big\}
\]

and $d(j, i) \defeq \text{min}\{B_{i, j}(\tau) \ne 0\}$ if $j \in \pa{i}$.

\begin{lemma}
  \label{lem:pwgc_anc}
  $x_j \pwgc x_i$ if and only if at least one of the following hold

  \begin{enumerate}
    \item{$j \in \anc{i}$}
    \item{$\anc{i}\cap\anc{j} \ne \emptyset$.} %and $d(k, j) < d(k, i)$}
  \end{enumerate}
\end{lemma}

% \begin{remark}
%   This lemma formalizes the intuition of \textit{information flow} in $\gcg$.  We have $x_j \pwgc x_i$ whenever $i$ is descended from $j$, or if they have a common ancestor (call it $k$) and ``information'' from $k$ reaches $j$ before it reaches $i$.
% \end{remark}

\begin{proof}
\end{proof}

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

\begin{theorem}[Tree Recovery]
  \label{thm:tree_recovery}
  If the Granger-Causality graph $\gcg$ for $x(t)$ is a
  directed forest, then $\gcg$ can be inferred from pairwise
  causality tests alone.
\end{theorem}

\section{Structure Learning}
\subsection{Local Search Heuristics}

\subsection{Edge-Wise Grouped LASSO}
Minimize the sparsity promoting regularized least squares problem and choose the hyper-parameters against the Akaike information criteria.

\begin{equation}
  L(\lambda, p) \defeq \underset{B}{\text{minimize}}\ \frac{1}{T}\sum_{t = 1}^T||x(t) - \sum_{\tau = 1}^pB(\tau)x(t - \tau)||_2^2 + \lambda \sum_{i, j}\big[\alpha||B_{i, j}||_2 + (1 - \alpha)||B_{i, j}||_1\big]
\end{equation}

\begin{equation}
  \underset{\lambda, p}{\text{minimize}}\ L(\lambda, p) + \mathsf{AIC}(B^{(\lambda, p)})
\end{equation}

\subsection{Bayesian Posterior Thresholding}
Consider the Bayesian model
